---
title: "Overweightness and SES"
author: "Norah Aalders 2889433, Sinem Göral 2817932"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment
text

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages()
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(readxl)
```

# Title Page

Include your names

Group 2.6

Tutorial Lecturer: Fatima Amankour

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   Overweightness in the Netherlands: A Social Problem Hidden in Plain Sight

Walk through any Dutch city today and the numbers get to ignore. More than half of Dutch adults are overweight, one in six has obesity (Rijksinstituut voor Volksgezondheid en Milieu [RIVM], 2026), and if current trends continue, two in three people could be affected by 2050. What was once seen as a personal health issue has become a major social challenge.
This became clear when Zorginstituut Nederland announced that bariatric surgery will now be covered by basic health insurance for teenagers aged 13 to 18 with severe obesity. Although only a small number of young people will qualify, the decision reflects growing recognition that lifestyle interventions alone are not always enough.
To understand obesity, however, we must look beyond individual choices. In the Netherlands, obesity is strongly linked to social inequality. People with lower levels of education are significantly more likely to be overweight than university graduates (RIVM, 2026). Lower-income groups often have less access to healthy food, safe places to exercise, and supportive living environments. Research also shows that factors such as neighbourhood safety, green spaces, and social cohesion influence obesity rates.(OECD & European Observatory on Health Systems and Policies, 2025).
The consequences extend beyond health. Obesity is associated with higher healthcare costs, lower earnings, increased sick leave, and reduced economic productivity. Children with obesity are more likely to face educational, social, and health disadvantages later in life.
This argues that obesity in the Netherlands should be viewed as a social problem rather than simply a matter of personal responsibility. Addressing it effectively requires tackling the structural inequalities that make obesity more common among disadvantaged groups.





# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
gewicht_regios <- read_xlsx("data/Gezondheidsmonitor__regio__2024_04062026_114221.xlsx")
saveRDS(gewicht_regios, "data/gewicht_regios.rds")

overgewicht_over_tijd <- read_xlsx("data/Lengte__onder__en_overgewicht_vanaf_1981_04062026_120636.xlsx")
saveRDS(overgewicht_over_tijd, "data/overgewicht_over_tijd.rds")

inkomen_dataset <- read_xlsx("data/Inkomens_Huishouden.xlsx")
saveRDS(inkomen_dataset, "data/inkomen_dataset.rds")

opleiding_dataset <- read_xlsx("data/Onderwijs_per_provincie.xlsx")
saveRDS(opleiding_dataset, "data/opleiding_dataset.rds")


```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}



```


https://www.staatvenz.nl/kerncijfers/overgewicht-en-obesitas-volwassenen

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning_1}
overgewicht_1981 = overgewicht_over_tijd[1,5]
overgewicht_2021 = overgewicht_over_tijd[10,5]

```

```{r data_cleaning_2}
time_all_ages <- overgewicht_over_tijd[[4]][1:10]
perc_overgewicht <- overgewicht_over_tijd[[5]][1:10]
perc_obesitas <- overgewicht_over_tijd[[7]][1:10]
```

```{r data_cleaning_3}
opleiding_provincie <- opleiding_dataset[c(1, 6:17),]

hoogopgeleide_provincie <- data.frame(Regio = opleiding_provincie[[4]],
                          Percentage_hoogopgeleide = opleiding_provincie[[9]] + opleiding_provincie[[10]])

middelopgeleide_provincie <- data.frame(Regio = opleiding_provincie[[4]],
                            Percentage_middelopgeleide = opleiding_provincie[[8]])


laagopgeleide_provincie <- data.frame(Regio = opleiding_provincie[[4]],
                          Percentage_laagopgeleide = opleiding_provincie[[6]] + opleiding_provincie[[7]])

```

```{r data_cleaning_4}
inkomen_provincie <- inkomen_dataset[1:13,]
gemiddeld_inkomen <- data.frame(Regio = inkomen_provincie[[4]],
                                "Gemiddeld_inkomen(x 1000)" = inkomen_provincie[[6]])
```

```{r data_cleaning_4}
data_regios <- gewicht_regios[grepl("\\(GG\\)", gewicht_regios[[3]] ), 3:12]
```

The dataset (data_regios) is built up in regions instead of provences. To compare this dataset to the other datasets, which are built up in provences, the regions need to be transformed into provences aswell. This is done with the following data from AlleCijfers.nl (AlleCijfers, 2026).

```{r data_cleaning_function}

data_analysis <- function(regio1 = 0, data1 = 0, regio2 = 0, data2 = 0, regio3 = 0, data3 = 0, regio4 = 0, data4 = 0, regio5 = 0,  data5 = 0){
  totaal_provincie <- regio1 + regio2 + regio3 + regio4
  data_provincie <- ((regio1 / totaal_provincie) * data1) + ((regio2 / totaal_provincie) * data2) + ((regio3 / totaal_provincie) * data3) + ((regio4 / totaal_provincie) * data4)  
  
  return(data_provincie)
  
}

```

```{r data_GGD_regions(AlleCijfers, 2026)}
groningen <- 601286

friesland <- 665208

drenthe <- 508109

ijsselland <- 555879
twente <- 646496

flevoland <- 462816

noord_oost_gelderland <- 851876
gelderland_midden <- 731773
gelderland_zuid <- 589699

utrecht <- 1415104

hollands_noorden <- 687079
zaanstreek <- 354021
kennemerland <- 570752
amsterdam <- 1151282
gooi <- 246252

haaglanden <- 1176111
hollands_midden <- 854306
rotterdam <- 1376801
zuid_holland_zuid <- 475490

zeeland <- 394538

west_brabant <-734825
hart_brabant <- 1122960
brabant_zuidoost <- 819169

limburg_noord <- 537019
limburg_zuid <- 601355

nederland <- groningen + friesland + drenthe + ijsselland + twente +flevoland + noord_oost_gelderland + gelderland_midden + gelderland_zuid + utrecht + hollands_noorden + zaanstreek + kennemerland +amsterdam + gooi + haaglanden + hollands_midden + rotterdam + zuid_holland_zuid + zeeland + west_brabant + hart_brabant + brabant_zuidoost + limburg_noord + limburg_zuid
```

```{r}
gezondheids_data_provincies <- rbind(Nederland = data_nederland <- data_analysis(nederland, gewicht_regios[1, 4:12]),
                              Groningen = data_groningen <- data_analysis(groningen, data_regios[1, 2:10]),

                              Friesland = data_friesland <- data_analysis(friesland, data_regios[24, 2:10]),

                              Drenthe = data_drenthe <- data_analysis(drenthe , data_regios[2, 2:10]),

                              Overrijsel = data_overrijsel <- data_analysis(ijsselland, data_regios[3,2:10], twente, data_regios[4,2:10]),

                              Flevoland = data_flevoland <- data_analysis(flevoland , data_regios[8, 2:10]),
      
                              Gelderland = data_gelderland <- data_analysis(noord_oost_gelderland, data_regios[5, 2:10], gelderland_midden, data_regios[6, 2:10], gelderland_zuid, data_regios[7, 2:10]),
                              
                              Utrecht = data_utrecht <- data_analysis(utrecht, data_regios[9, 2:10]),
                              
                              Noord_Holland = data_noord_holland <- data_analysis(hollands_noorden , data_regios[10, 2:10], zaanstreek, data_regios[25, 2:10], kennemerland, data_regios[11, 2:10], amsterdam, data_regios[12, 2:10], gooi, data_regios[13, 2:10]),

                              Zuid_Holland = data_zuid_holland <- data_analysis(haaglanden, data_regios[23, 2:10], hollands_midden, data_regios[14, 2:10], rotterdam, data_regios[15, 2:10], zuid_holland_zuid, data_regios[16, 2:10]),

                              Zeeland = data_zeeland <- data_analysis(zeeland , data_regios[17, 2:10]),

                              Brabant = data_brabant <- data_analysis(west_brabant, data_regios[18, 2:10], hart_brabant, data_regios[19, 2:10], brabant_zuidoost, data_regios[20, 2:10]),

                              Limburg = data_limburg <- data_analysis(limburg_noord, data_regios[21, 2:10], limburg_zuid, data_regios[22, 2:10]))

gezondheids_data_provincies <- as.data.frame(gezondheids_data_provincies)
```

```{r}
Overgewicht_provincies <- gezondheids_data_provincies[c(3, 4)]
```


Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
proc_ver_overgewicht = ((overgewicht_2021 - overgewicht_1981) / overgewicht_1981) * 100
```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}
ggplot() +
  geom_line(aes(x = time_all_ages,
                y = perc_overgewicht,
                color = "Overgewicht")) +
  geom_line(aes(x = time_all_ages,
                y = perc_obesitas,
                color = "Obesitas")) +
    geom_point(aes(x = time_all_ages,
                 y = perc_overgewicht)) +
  geom_point(aes(x = time_all_ages,
                 y = perc_obesitas)) +
  ylim(0,50) +
  labs(title = "Percentage of overweight people over time in the Netherlands",
       x = "Year",
       y = "Percentage of people")

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
AlleCijfers. (2026, April). GGD-regio's: overzicht en statistieken per GGD-regio.https://allecijfers.nl/ggd-regios/

Centraal Bureau voor de Statistiek. (2024, oktober). Obesitas afgelopen 40 jaar verdrievoudigd. https://www.cbs.nl/nl-nl/nieuws/2024/10/obesitas-afgelopen-40-jaar-verdrievoudigd

GGD Zaanstreek-Waterland. (2022). Wat kost overgewicht? https://www.ggdzw.nl/wat-kost-overgewicht/

NOS. (2026, 4 juni). Maagverkleining voor kinderen met obesitas in basispakket als laatste redmiddel. https://nos.nl/artikel/2617067-maagverkleining-voor-kinderen-met-obesitas-in-basispakket-als-laatste-redmiddel

OECD & European Observatory on Health Systems and Policies. (2025). State of health in the EU: Netherlands — Country health profile 2025. OECD Publishing. https://www.oecd.org/en/publications/country-health-profile-2025-country-notes_c0e8b8be-en/netherlands_95a85df5-en.html

Rijksinstituut voor Volksgezondheid en Milieu. (2026). Overgewicht — Volwassenen. Volksgezondheidenzorg.info. https://www.vzinfo.nl/overgewicht/volwassenen

Rijksinstituut voor Volksgezondheid en Milieu. (2026). Sociaaleconomische gezondheidsverschillen. Volksgezondheidenzorg.info. https://www.vzinfo.nl/sociaaleconomische-gezondheidsverschillen

