---
title: "Overweightness and SES"
author: "Norah Aalders 2889433, Sinem Göral 2817932"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

text

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("yaml")
#install.packages("rmarkdown")
#install.packages()
#install.packages("sf")   # alleen één keer nodig

```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(readxl)
library(sf)
```

# Title Page

Include your names
David Hers, 2905659
Norah Aalders, 2889433
Casper Flipse, 2898924
Sinem Göral, 2817932
Chris Kool, 2898107
Roel van Soelen, 2898917

Group 2.6

Tutorial Lecturer: Fatima Amankour

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   Overweightness in the Netherlands: A Social Problem Hidden in Plain Sight

Walk through any Dutch city today and the numbers get to ignore. More than half of Dutch adults are overweight, one in six has obesity (Rijksinstituut voor Volksgezondheid en Milieu [RIVM], 2026), and if current trends continue, two in three people could be affected by 2050. What was once seen as a personal health issue has become a major social challenge.
This became clear when Zorginstituut Nederland announced that bariatric surgery will now be covered by basic health insurance for teenagers aged 13 to 18 with severe obesity. Although only a small number of young people will qualify, the decision reflects growing recognition that lifestyle interventions alone are not always enough.
To understand obesity, however, we must look beyond individual choices. In the Netherlands, obesity is strongly linked to social inequality. People with lower levels of education are significantly more likely to be overweight than university graduates (RIVM, 2026). Lower-income groups often have less access to healthy food, safe places to exercise, and supportive living environments. Research also shows that factors such as neighbourhood safety, green spaces, and social cohesion influence obesity rates.(OECD & European Observatory on Health Systems and Policies, 2025).
The consequences extend beyond health. Obesity is associated with higher healthcare costs, lower earnings, increased sick leave, and reduced economic productivity. Children with obesity are more likely to face educational, social, and health disadvantages later in life.
This argues that obesity in the Netherlands should be viewed as a social problem rather than simply a matter of personal responsibility. Addressing it effectively requires tackling the structural inequalities that make obesity more common among disadvantaged groups.



- Overweightness in the Netherlands: A Social Problem Hidden in Plain Sight

Walk through any Dutch city today and the numbers get to ignore. More than half of Dutch adults are overweight, one in six has obesity (Rijksinstituut voor Volksgezondheid en Milieu [RIVM], 2026), and if current trends continue, two in three people could be affected by 2050. What was once seen as a personal health issue has become a major social challenge. This became clear when Zorginstituut Nederland announced that bariatric surgery will now be covered by basic health insurance for teenagers aged 13 to 18 with severe obesity. Although only a small number of young people will qualify, the decision reflects growing recognition that lifestyle interventions alone are not always enough. To understand obesity, however, we must look beyond individual choices. In the Netherlands, obesity is strongly linked to social inequality. People with lower levels of education are significantly more likely to be overweight than university graduates (RIVM, 2026). Lower-income groups often have less access to healthy food, safe places to exercise, and supportive living environments. Research also shows that factors such as neighbourhood safety, green spaces, and social cohesion influence obesity rates.(OECD & European Observatory on Health Systems and Policies, 2025). The consequences extend beyond health. Obesity is associated with higher healthcare costs, lower earnings, increased sick leave, and reduced economic productivity. Children with obesity are more likely to face educational, social, and health disadvantages later in life. This argues that obesity in the Netherlands should be viewed as a social problem rather than simply a matter of personal responsibility. Addressing it effectively requires tackling the structural inequalities that make obesity more common among disadvantaged groups.

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
gewicht_regios <- read_xlsx("data/Gezondheidsmonitor__regio__2024_04062026_114221.xlsx")
saveRDS(gewicht_regios, "data/gewicht_regios.rds")

overgewicht_over_tijd <- read_xlsx("data/Lengte__onder__en_overgewicht_vanaf_1981_04062026_120636.xlsx")
saveRDS(overgewicht_over_tijd, "data/overgewicht_over_tijd.rds")

inkomen_dataset <- read_xlsx("data/Inkomens_Huishouden.xlsx")
saveRDS(inkomen_dataset, "data/inkomen_dataset.rds")

opleiding_dataset <- read_xlsx("data/Onderwijs_per_provincie.xlsx")
saveRDS(opleiding_dataset, "data/opleiding_dataset.rds")


```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}



```

``` r
inline_code = TRUE
```
In this project, we utilize two datasets released by CBS (Statistics Netherlands) that collectively offer insight into the health of the Dutch population. The initial dataset, Gezondheidsmonitor; adults 18 years and older, region, 2024, centers on individuals aged 18 and above residing in the Netherlands. CBS compiled it in partnership with the National Institute for Public Health and the Environment (RIVM) and the local Public Health Services (GGDs). The data collection provides insights into health conditions, lifestyle choices, and social elements across various geographical tiers, such as municipalities, GGD regions, provinces, and at the national level. Essential indicators consist of self-assessed health, rates of chronic diseases, overweight and obesity, levels of physical exercise, involvement in sports, tobacco use, alcohol intake, and informal caregiving. The information is primarily shown as percentages of the adult demographic and offers a comprehensive regional overview of health and wellness in 2024.The second dataset, Lengte en gewicht van personen, ondergewicht en overgewicht; vanaf 1981, provides a longitudinal view of body weight and obesity trends in the Netherlands. This dataset includes individuals aged four and older and monitors trends from 1981 to 2025. It contains data regarding body mass index (BMI), moderate overweight, obesity, and underweight, categorized by age group, gender, and year. The data relies on self-reported weight and height, which CBS utilizes to determine BMI categories. In contrast to the initial dataset that highlights regional variations for one year, this dataset offers a national time series enabling researchers to explore changes in overweight and obesity across more than forty years.

Collectively, these datasets enhance each other effectively. The Health Monitor dataset offers an in-depth view of regional health disparities in 2024, whereas the height and weight dataset uncovers enduring national patterns of overweight and obesity. Integrating information from both sources allows for a clearer comprehension of obesity's prevalence in the Netherlands, its regional variations, and its historical trends.

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning_1}
overgewicht_1981 = overgewicht_over_tijd[1,5]
overgewicht_2021 = overgewicht_over_tijd[10,5]
```

```{r data_cleaning_2}
perc_overgewicht <- full_join(overgewicht_over_tijd[1:13, c(4,5,7)], full_join(overgewicht_over_tijd[40:52, c(4,5)], overgewicht_over_tijd[79:91, c(4,5)], by = "Perioden"), by = "Perioden")
```

```{r data_cleaning_3}
opleiding_provincie <- data.frame("Regio" = opleiding_dataset[[4]][6:17],
                                  "Percentage_laagopgeleide" = opleiding_dataset[[6]][6:17] + opleiding_dataset[[7]][6:17], 
                                  "Percentage_middelopgeleide" = opleiding_dataset[[8]][6:17],
                                  
                                  "Percentage_hoogopgeleide" = opleiding_dataset[[9]][6:17] + opleiding_dataset[[10]][6:17]
)

```

```{r data_cleaning_4}
gemiddeld_inkomen <- data.frame(Regio = inkomen_dataset[[4]][2:13],
                                "Gemiddeld_inkomen(x 1000)" = inkomen_dataset[[6]][2:13])
```

```{r data_cleaning_4}
data_regios <- gewicht_regios[grepl("\\(GG\\)", gewicht_regios[[3]] ), 3:12]
```

The dataset (data_regios) is built up in regions instead of provences. To compare this dataset to the other datasets, which are built up in provences, the regions need to be transformed into provences aswell. This is done with the following data from AlleCijfers.nl (AlleCijfers, 2026).

```{r data_cleaning_function}

data_analysis <- function(regio1 = 0, data1 = 0, regio2 = 0, data2 = 0, regio3 = 0, data3 = 0, regio4 = 0, data4 = 0, regio5 = 0,  data5 = 0){
  totaal_provincie <- regio1 + regio2 + regio3 + regio4
  data_provincie <- ((regio1 / totaal_provincie) * data1) + ((regio2 / totaal_provincie) * data2) + ((regio3 / totaal_provincie) * data3) + ((regio4 / totaal_provincie) * data4)  
  
  return(data_provincie)
  
}

```

```{r data_GGD_regions(AlleCijfers, 2026)}
groningen <- 601286

friesland <- 665208

drenthe <- 508109

ijsselland <- 555879
twente <- 646496

flevoland <- 462816

noord_oost_gelderland <- 851876
gelderland_midden <- 731773
gelderland_zuid <- 589699

utrecht <- 1415104

hollands_noorden <- 687079
zaanstreek <- 354021
kennemerland <- 570752
amsterdam <- 1151282
gooi <- 246252

haaglanden <- 1176111
hollands_midden <- 854306
rotterdam <- 1376801
zuid_holland_zuid <- 475490

zeeland <- 394538

west_brabant <-734825
hart_brabant <- 1122960
brabant_zuidoost <- 819169

limburg_noord <- 537019
limburg_zuid <- 601355

```

```{r}
gezondheids_data_provincies <- rbind(Groningen  = data_analysis(groningen, data_regios[1, 2:10]),

                              Fryslân = data_analysis(friesland, data_regios[24, 2:10]),

                              Drenthe = data_analysis(drenthe , data_regios[2, 2:10]),

                              Overijssel = data_analysis(ijsselland, data_regios[3,2:10], twente, data_regios[4,2:10]),

                              Flevoland = data_analysis(flevoland , data_regios[8, 2:10]),
      
                              Gelderland = data_analysis(noord_oost_gelderland, data_regios[5, 2:10], gelderland_midden, data_regios[6, 2:10], gelderland_zuid, data_regios[7, 2:10]),
                              
                              Utrecht = data_analysis(utrecht, data_regios[9, 2:10]),
                              
                              "Noord-Holland" = data_analysis(hollands_noorden , data_regios[10, 2:10], zaanstreek, data_regios[25, 2:10], kennemerland, data_regios[11, 2:10], amsterdam, data_regios[12, 2:10], gooi, data_regios[13, 2:10]),

                              "Zuid-Holland" = data_analysis(haaglanden, data_regios[23, 2:10], hollands_midden, data_regios[14, 2:10], rotterdam, data_regios[15, 2:10], zuid_holland_zuid, data_regios[16, 2:10]),

                              Zeeland = data_analysis(zeeland , data_regios[17, 2:10]),

                              "Noord-Brabant" = data_analysis(west_brabant, data_regios[18, 2:10], hart_brabant, data_regios[19, 2:10], brabant_zuidoost, data_regios[20, 2:10]),

                              Limburg = data_analysis(limburg_noord, data_regios[21, 2:10], limburg_zuid, data_regios[22, 2:10]))

gezondheids_data_provincies <- as.data.frame(gezondheids_data_provincies)

gezondheids_data_provincies$Provincie <- rownames(gezondheids_data_provincies)
rownames(gezondheids_data_provincies) <- NULL
gezondheids_data_provincies <- gezondheids_data_provincies[, c("Provincie",
                                                               setdiff(names(gezondheids_data_provincies), "Provincie"))]
```

```{r}
names(gezondheids_data_provincies)[4] <- "Overgewicht"
names(gezondheids_data_provincies)[names(gezondheids_data_provincies) == "Provincie"] <- "Regio"
```

```{r data_mergen}
name_change <- full_join(opleiding_provincie, gemiddeld_inkomen)

```

```{r}
name_change[1,1] <- "Groningen"
name_change[2,1] <- "Fryslân"
name_change[3,1] <- "Drenthe"
name_change[4,1] <- "Overijssel"
name_change[5,1] <- "Flevoland"
name_change[6,1] <- "Gelderland"
name_change[7,1] <- "Utrecht"
name_change[8,1] <- "Noord-Holland"
name_change[9,1] <- "Zuid-Holland"
name_change[10,1] <- "Zeeland"
name_change[11,1] <- "Noord-Brabant"
name_change[12,1] <- "Limburg"
```

```{r}
dataset_map <- left_join(gezondheids_data_provincies[c(1, 4)], name_change)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
proc_ver_overgewicht = ((overgewicht_2021 - overgewicht_1981) / overgewicht_1981) * 100
```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}
perc_over_time_plot = ggplot() +
  geom_line(aes(x = perc_overgewicht[[1]],
                y = perc_overgewicht[[2]],
                color = "Overweight")) +
  geom_line(aes(x = perc_overgewicht[[1]],
                y = perc_overgewicht[[3]],
                color = "Obesity")) +
    geom_point(aes(x = perc_overgewicht[[1]],
                 y = perc_overgewicht[[2]])) +
  geom_point(aes(x = perc_overgewicht[[1]],
                 y = perc_overgewicht[[3]])) +
  scale_x_continuous(breaks = seq(1980, 2025, by = 5)) +
  ylim(0,50) +
  labs(title = "Percentage of overweight people\n over time in the Netherlands (1981-2025)",
       x = "Year",
       y = "Percentage of people") +
      theme(plot.title = element_text(hjust = 0.5))

ggsave("Graphs/Percentage_overgewicht_over_tijd.png", perc_over_time_plot, width = 9, height = 6)
```

As shown in the graph, the percentage of overweight/obese people has increased at a relatively constant rate over time, with the exception of a larger increase between 1995 and 2000.

## 3.4 Visualize spatial variation

```{r visualise_map}
provincies_kaart <- st_read(paste0(
  "https://service.pdok.nl/cbs/gebiedsindelingen/2024/wfs/v1_0?",
  "request=GetFeature&service=WFS&version=2.0.0",
  "&typeName=provincie_gegeneraliseerd&outputFormat=application/json"
))

provincies_kaart$statnaam 

nl_map <- provincies_kaart |> left_join(dataset_map, by = c("statnaam" = "Regio")) 


map_overweight <- ggplot(nl_map) +
  geom_sf(aes(fill = Overgewicht ), color = "white", linewidth = 0.4) +
  scale_fill_gradient(name = "Percentage of \noverweight people \n(reversed)", high = "#e8f5e9", 
  low = "#1b5e20") +
  labs(
    title   = "Percentage of overweight people\n by province in the Netherlands (2024)"  ) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave("graphs/map_overweight_people.png", map_overweight, width = 6, height = 10)
```

```{r}
map_high_education <- ggplot(nl_map) +
  geom_sf(aes(fill = Percentage_hoogopgeleide ), color = "white", linewidth = 0.4) +
  scale_fill_gradient(name = "Percentage highly \neducated people", low = "#e8f5e9", high = "#1b5e20") +
  labs(
    title   = "Percentage of highly educated people \n by province in the Netherlands (2024)"  ) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave("Graphs/map_high-education.png", map_high_education, width = 6, height = 10)
```

```{r}
map_lower_education <- ggplot(nl_map) +
  geom_sf(aes(fill = Percentage_laagopgeleide ), color = "white", linewidth = 0.4) +
  scale_fill_gradient(name = "Percentage lower\n educated people \n(reversed)", high = "#e8f5e9", low = "#1b5e20") +
  labs(
    title   = "Percentage of lower educated people\n by province in the Netherlands (2024)"  ) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave("Graphs/map_lower_education.png", map_lower_education, width = 6, height = 10)
```

```{r}
map_avg_income <- ggplot(nl_map) +
  geom_sf(aes(fill =  Gemiddeld_inkomen.x.1000.), color = "white", linewidth = 0.4) +
  scale_fill_gradient(name = "Average income \n(x1000)", low = "#e8f5e9", high = "#1b5e20") +
  labs(
    title   = "Average income by province\n in the Netherlands"  ) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave("Graphs/map_avg_income.png", map_avg_income, width = 6, height = 10)
```

As shown in the map representing hte perecentage of overweight and obese individuals, the provinces Utrecht, Noord-Holland, and Groningen appear to have the lowest prevalence rates. In contrast, Flevoland, Limburg and Drenthe have the highest percentage of overweight and obese residents. Furthermore a possible correlation can be found between overweight and obesity and educational level. The provinces with the lowest prevalence rates also have the highest level of education, whereas Flevoland and Drenthe, which have a relatively lower education level, show a higher percentage of overweight and obese residents. The level of income and overweight appear to have a weaker relationship. 


## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}

plot_men_vs_women <- ggplot() +
  geom_line(aes(x = perc_overgewicht[[1]],
                y = perc_overgewicht[[2]],
                color = "Overweight")) +
  geom_line(aes(x = perc_overgewicht[[1]],
                y = perc_overgewicht[[4]],
                color = "Men")) +
  geom_line(aes(x = perc_overgewicht[[1]],
                y = perc_overgewicht[[5]],
                color = "Women")) +
  geom_point(aes(x = perc_overgewicht[[1]],
                y = perc_overgewicht[[2]])) +
  geom_point(aes(x = perc_overgewicht[[1]],
                 y = perc_overgewicht[[4]])) +
  geom_point(aes(x= perc_overgewicht[[1]],
              y = perc_overgewicht[[5]]))  +
  scale_color_manual( values = c("Overweight" = "black",
                                 "Men" = "blue",
                                 "Women" = "deeppink")) +
  scale_x_continuous(breaks = seq(1980, 2025, by = 5)) +
  ylim(0,50) +
  labs(title = "Percentage of overweight men \ncompared to overweight woman over time (1981-2025)",
       x = "Year",
       y = "Percentage of people",
       color = "group") +
    theme(plot.title = element_text(hjust = 0.5))
  
ggsave("Graphs/plot_men_vs_women.png", plot_men_vs_women, width = 9, height = 6)
```
Over the 44-year period shown, the percentage of overweight adults has always been higher among men than among women. Although this gap has shrunk in recent years, men still show a higher rates of overweight. furthermore, the percentage of overweight men and the percentage of overweight women have generally moved together over time. 


## 3.6 Event analysis

in 2005, the Dutch covenant overweight was introduced and signed by several organizations with the goal of preventing overweight and obesity. Here it is tested if the percentage of overweight and obese people has reduced since its introduction.

```{r analysis}
plot_event <- ggplot() +
  geom_line(aes(x = perc_overgewicht[[1]],
                y = perc_overgewicht[[2]],
                color = "Overgewicht")) +
  geom_line(aes(x = perc_overgewicht[[1]],
                y = perc_overgewicht[[3]],
                color = "Obesitas")) +
    geom_point(aes(x = perc_overgewicht[[1]],
                 y = perc_overgewicht[[2]])) +
  geom_point(aes(x = perc_overgewicht[[1]],
                 y = perc_overgewicht[[3]])) +
  geom_vline(xintercept = 2005, linetype = "dashed")+
  annotate("text", x = 2012, y = 30, size = 3, label = "Covenant overweight") +
  scale_x_continuous(breaks = seq(1980, 2025, by = 5)) +
  ylim(0,50) +
  labs(title = "Percentage of overweight people\n over time in the Netherlands(1981-2025)",
       x = "Year",
       y = "Percentage of people") +
    theme(plot.title = element_text(hjust = 0.5))
  
ggsave("Graphs/®plot_event.png", plot_event, width = 9, height = 6)

```

As shown in the temporal variation, the percentage of overweight/obese people has increased at a relatively constant rate over time. This rate does not appear to have changed from 2005 onwards. Consequently, no clear change can be observed following the introduction of the covenant on overweight and obesity, however based on the graph alone, it is not possible to determine whether the covenant had an effect. 

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
AlleCijfers. (2026, April). GGD-regio's: overzicht en statistieken per GGD-regio.https://allecijfers.nl/ggd-regios/

Centraal Bureau voor de Statistiek. (2024, oktober). Obesitas afgelopen 40 jaar verdrievoudigd. https://www.cbs.nl/nl-nl/nieuws/2024/10/obesitas-afgelopen-40-jaar-verdrievoudigd

GGD Zaanstreek-Waterland. (2022). Wat kost overgewicht? https://www.ggdzw.nl/wat-kost-overgewicht/

NOS. (2026, 4 juni). Maagverkleining voor kinderen met obesitas in basispakket als laatste redmiddel. https://nos.nl/artikel/2617067-maagverkleining-voor-kinderen-met-obesitas-in-basispakket-als-laatste-redmiddel

OECD & European Observatory on Health Systems and Policies. (2025). State of health in the EU: Netherlands — Country health profile 2025. OECD Publishing. https://www.oecd.org/en/publications/country-health-profile-2025-country-notes_c0e8b8be-en/netherlands_95a85df5-en.html

Rijksinstituut voor Volksgezondheid en Milieu. (2026). Overgewicht — Volwassenen. Volksgezondheidenzorg.info. https://www.vzinfo.nl/overgewicht/volwassenen

Rijksinstituut voor Volksgezondheid en Milieu. (2026). Sociaaleconomische gezondheidsverschillen. Volksgezondheidenzorg.info. https://www.vzinfo.nl/sociaaleconomische-gezondheidsverschillen

